### 2018-07-24 chapters/03-translator/forward-trans.tex
Update until forward/backword translation
Finalize first version of the journal paper Still 1.000 words too long

\section{The Forward Translation}\label{sec:forward-translation}
%\glsreset{teo}
The translation process starts with semantic \LaTeX{} expressions. These expressions get analyzed and parsed by the \gls{mlp}, which produces a \gls{mlp-pt} of the input expression with additional information about each symbol. An abstract translator class analyzes each node and delegates them to specialized subtranslators. In this process an object called \gls{teo} is build. This \gls{teo} is needed to rebuild a string representation of the mathematical expression given by the \gls{mlp-pt}.

The overall forward translation process is explained in figure~\ref{fig:forward-trans}. All translation patterns and related information are stored in the DLMF/DRMF tables. These tables are converted by the \verb|lexicon-creator.jar| to the \verb|DLMF-macros-lexicon.txt| lexicon file. Together with the \verb|global-lexicon.txt| file, the \gls{mlp-pt} will be created by the \gls{mlp}. The \verb|latex-converter.jar| takes a string representation of a semantic \LaTeX{} expression and uses the \gls{mlp} as well as our \verb|Translator| to create an appropriate string representation for a specified \gls{cas}. The following \cref{subsec:analyze-mlp,subsec:delegation,subsec:translated-expr,subsec:add-info} will focus on the \verb|Translator|.

\begin{figure}[ht]
	\vspace{-10pt}
	\centering
	\includegraphics[clip, trim=0.2cm 0.2cm 0.2cm 0.2cm, scale=0.72]{ForwardTranslationProcess.pdf}
	\caption{Process diagram of a forward translation process. The MLP generates the PT based on lexicon and JSON files. The PT will be translated to different CAS.}
	\label{fig:forward-trans}
	\vspace{-10pt}
\end{figure}

\subsection{Analyzing the MLP-Parse Tree}\label{subsec:analyze-mlp}
The main task is to analyze the \gls{mlp-pt}. Remember that the \gls{mlp-pt} is different to expression trees as described in section~\ref{subsec:pom-mlp}.

Since the \gls{bnf} does not define rules for semantic macros, each argument of the semantic macro and each $@$ symbol are following siblings of the semantic macro node. That is the reason, why we stored the number of parameters, variables and $@$ symbols in the lexicon files. Otherwise, the translator could not find the end of a semantic macro in the \gls{mlp-pt}.

Figure~\ref{fig:syntax-tree-usecase} visualizes the \gls{mlp-pt} of the Jacobi polynomial example from table~\ref{tab:JacobiP-usecase}. Because of these differences to expression trees, a backward conversion of the \gls{mlp-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls{teo}. Of course, the \gls{teo} is deeply incorporated into the translation process, but in this subsection we will focus on the general idea and process of the translation. We will focus on the usage of \gls{teo} in subsection~\ref{subsec:translated-expr}.

\begin{figure}[ht]
	\centering
	\includegraphics[clip, trim=0.2cm 0.2cm 0.2cm 0.2cm, scale=0.75]{SyntaxTreeUseCase.pdf}
	\caption{MLP-PT for a Jacobi polynomial using the DLMF/DRMF \LaTeX{} macro. Each leaf contains information from the lexicon files.}
	\label{fig:syntax-tree-usecase}
\end{figure}

The general idea of a translation of a \gls{mlp-pt} is a simple recursive algorithm, such as algorithm~\ref{alg:simple-translation}. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively.

\begin{wrapfigure}{r}{0.5\textwidth}
\vspace{-20pt}
\begin{minipage}{0.5\textwidth}
\begin{algorithm}[H]
\caption{Simple translation algorithm for the MLP-Parse Tree}\label{alg:simple-translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of the parse tree $T$
	\Procedure{translate}{$r$}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$);
	\Else
		\ForAll{children $v_n$ of $r$}
			\State {\scriptsize TRANSLATE}($v_n$);
		\EndFor
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\end{minipage}
\vspace{-14pt}
\end{wrapfigure}

But this approach is not completely feasible, because sometimes the algorithm needs to look ahead and check the following siblings for a valid translation, for example in case of a semantic macro with arguments (see figure~\ref{fig:syntax-tree-usecase}). Therefore, the {\footnotesize \verb|TRANSLATE_LEAF|} function also needs to know the following siblings of the current node. To avoid multiple translations of the same node, a list representation would be a better solution. Algorithm~\ref{alg:translation} explains an abstract version of the final algorithm for our translation tool.

If the root $r$ is a leaf, it can be translated as a leaf. Eventually, some of the following siblings are needed in order to translate $r$ correctly. If $r$ is not a leaf, it contains one or more children. Therefore, we can call the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively for the children. Once we have translated $r$, we can go a step further and translate the next node. Line~\ref{line:ifNotEmpty} checks if there are following siblings left and calls the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively in that case.

\begin{algorithm}[ht]
\caption{Abstract translation algorithm to translate MLP-Parse trees.}\label{alg:translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of a MLP-Parse tree $T$. List \textit{following\_siblings} with the following siblings of $r$. The list can be empty.
	\Procedure{abstract\_translator}{$r$, \textit{following\_siblings}}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$, \textit{following\_siblings\_of\_r});\label{line:transLeaf}
	\Else
		\State \textit{siblings} $ = r$.getChildren(); \Comment{\textit{siblings} is a list of children}
		\State {\scriptsize ABSTRACT\_TRANSLATOR}(\textit{siblings}.removeFirst(), \textit{siblings});\label{line:transSeq}
	\EndIf
	\If{\textit{following\_siblings} is not empty}\label{line:ifNotEmpty}
		\State $r =$ \textit{following\_siblings}.removeFirst();
		\State {\scriptsize ABSTRACT\_TRANSLATOR}($r$, \textit{following\_siblings});
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}

With this approach we are able to translate most of the semantic \LaTeX{} expressions. To take a look at the problems with this approach, we need to think about the operators and their standard notations in mathematics. There are many types of notation used to represent formulae. For example, the \gls{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between their operands, which only makes sense as long as the operator is a binary operator. For example, table~\ref{tab:notations} shows an expression in different notations.

\begin{table}[!ht]
\centering
\begin{tabular}{lc}
	\hline
	Notation & Expression \\
	\hline
	Infix & $(a+b) \cdot x$\\
	Prefix & $\cdot + a\ b\ x$\\
	Postfix & $a\ b + x\ \cdot$\\
	Functional & $\cdot(+(a, b), x)$\\
	\hline
\end{tabular}
\caption{The mathematical expression '$(a+b) \cdot x$' in infix, prefix, postfix and functional notation.}
\label{tab:notations}
\end{table}

In mathematical expressions, notations are mostly mixed, depending on the case and number of operands. For example, infix notation is common for binary operators ($+$, $-$, $\cdot$, $\mod$ etc.), while functional notations are obviously used for any kind of functions ($sin$, $cos$, etc.). Sometimes the same symbol is used in different notations to identify a different meaning. For example, the '$-$' as an unary operator is used in prefix notation to indicate the negative value of its operand, such as in '$-2$'. Of course, $-$ can also be the binary operator for subtraction, which is commonly used in infix notation. An example for the postfix notation is factorial, such as '$2!$'.

Most programming languages (and \gls{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, because it is easier to parse those notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls{mlp} has barely implemented mathematical grammatical rules, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from table~\ref{tab:notations} to four different \gls{mlp-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process because most \gls{cas} are familiar with the most common notations. Therefore, the translator does not need to know that $a$ and $b$ are the operands of the binary operator '$+$' in '$a+b$'. We could just translate the symbols in '$a+b$' in the same order as they appear in the expression and the \gls{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}%[label=\arabic*)]
\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to a \gls{cas} which uses a different notation.
\end{enumerate}

Problem~\ref{prob:1} should be obvious. Since we want to develop a translation tool and not a verification tool for mathematical \LaTeX{} expressions, we can assume syntactically correct input expressions and produce errors otherwise. Problem~\ref{prob:2} is more difficult to solve. If a user wants to support a \gls{cas} that uses prefix or postfix notation by default, the whole translator would fail in its current state. Supporting \gls{cas} with another notation would be a part of future work and will be mentioned in chapter~\ref{ch:conc-future-work}.

@@

In most cases, parentheses can be used to disambiguate expressions. We used them in table~\ref{tab:amb_ex} to clarify the different interpretations in example 2. But sometimes, even parentheses cannot solve a mistaken computation. For example, there is no way to add parentheses to force \Maple{} to compute $n!!$ as the double factorial function. Even $(n)!!$ will be interpreted as $(n!)!$. Rather than using the exclamation mark in \Maple, one could also use the functional notation. For example, the interpretations $(2!)!$ and $(2)!!$ can be distinguished in \Maple{} by using \verb|factorial(factorial(2))| and \verb|doublefactorial(2)| respectively.

As already mentioned, the structure of the \gls{mlp-pt} makes it difficult to change the notation for all kind of operators. Therefore, and especially because of the examples above, we change the notation during the forward translation process to the functional notation only for the factorial and double factorial function. Thus,
\begin{eqnarray*}
\verb|n!| &\overset{\langMaple}{\mapsto}& \verb|factorial(n)|,\\
\verb|n!!| &\overset{\langMaple}{\mapsto}& \verb|doublefactorial(n)|.
\end{eqnarray*}
The translator needs to presume some properties for the functions similar to \LaTeX, which only recognizes the very next element right after the caret for the superscript.

The biggest problem about translating the factorial (or double factorial) function is the common postfix notation for them. In this stage, algorithm~\ref{alg:translation} only translates the current node and if necessary the following siblings, but not the preceding siblings. Consider an expression such as '$4!$'. When the variable $r$ reaches the exclamation mark, the argument '$4$' of the function has already been translated. The current version of the algorithm has no access to this previously translated expression. Even more complicated, would be the case of the double factorial '$4!!$'. The following two approaches would solve this problem.

\begin{enumerate}
\item\label{appr:1} The translator always checks if the next two following siblings are one or two exclamation marks.
\item\label{appr:2} The current state of program has access to the previously translated expressions and can modify them retrospectively.
\end{enumerate}

Approach~\ref{appr:1} would be unnecessarily inefficient. Therefore we implement approach~\ref{appr:2}. That is another reason why we implement an extra object to organize translated expressions, the \gls{teo}. This object stores all parts of the previously translated expressions. The translator has access to these parts and is able to modify them afterwards. A more detailed explanation about the \gls{teo} follows in subsection~\ref{subsec:translated-expr}.
\newpage

\input{chapters/03-translator/all-types-table.tex}
\newpage

\subsection{Delegation of Translations}\label{subsec:delegation}

\begin{wrapfigure}{l}{0.45\textwidth}
	\vspace{-20pt}
	\centering
	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.4]{ForwardTranslator.pdf}
	\caption{A scheme of the forward translator and its specialized subtranslators.}
	\label{fig:ForwardDel}
	\vspace{-25pt}
\end{wrapfigure}

Algorithm~\ref{alg:translation} is a simplified version of the translator process. The main progress happens in \cref{line:transLeaf,line:transSeq}. There are several cases what $r$ can be. Table~\ref{tab:allTypesTable} on page \pageref{tab:allTypesTable} gives an overview of all different types in a \gls{mlp-pt}. A more detailed explanation of the types can be found in~\parencite{POM-Tagger}.

The \gls{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols, similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegates the translation to specialized subtranslators. This delegation process is implemented in \cref{line:transLeaf,line:transSeq} of algorithm~\ref{alg:translation}.

Figure~\ref{fig:ForwardDel} is a scheme of the forward translator and its subtranslators. The entry point for each translation process of the program is the \verb|SemanticLatexTranslator|. It delegates the translations for the expression through the \verb|AbstractTranslator| to the other subtranslators. The \verb|AbstractTranslator| is the interface of all other translators and coordinates the translation process by delegating translations for specific nodes to specialized classes. For example, the \verb|MacroTranslator| only translates semantic \LaTeX{} macros. The \verb|EmptyExpressionTranslator| is used to delegate the so called \textit{empty expressions} to other classes. An \textit{empty expression} is a node with children, but without a symbol itself, such as \textit{sequences} and \textit{balanced expressions}.

\subsubsection{Subtranslators}
The most interesting translators are the \verb|MacroTranslator| (translating semantic \LaTeX{} macros) and the \verb|SequenceTranslator| (translating all kinds of \textit{sequences}).

The \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls{mlp-pt}. If a node $n$ is a leaf and the represented symbol an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balances of parentheses in expressions. An expression such as '$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as '$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and an half-opened, half-closed interval. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. In addition, the parenthesis checker could simply be deactivated to allow mismatched parentheses in an expression. But this functionality is not implemented yet.

Another problem that the \verb|SequenceTranslator| solves is the position of multiplication signs in an expression. There are a couple of obvious choices to translate multiplication signs. For example, using \verb|\cdot| or \verb|\idot| will be obviously translated as a multiplication symbol. However, \verb|\idot| is fairly new and therefore not frequently used yet. Furthermore, the most common symbol for multiplications is still the white space or none symbol at all, as explained above. Consider the simple expression $2n\pi$. The \gls{mlp-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. Therefore, the \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if there should be a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as $f(x)$. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)|.

Besides the translation of sequences, the translation process of the \Macro s can be complex too; not only because of the structure of the \gls{mlp-pt}, but also because of the complex definition of the semantic macros. As explained in \ref{subsec:macros}, they can contain optional parameters, exponents are allowed before and after the arguments and the number of $@$ symbols varies. All these different cases will be solved by the \verb|MacroTranslator|.

Algorithm~\ref{alg:macro-translation} is the translation function of the \verb|MacroTranslator| without error handling. It analyzes the next element right after the macro first. As mentioned, this can be one of the following:
\begin{itemize}
\item an exponent such as '\verb|^2|'.
\item an optional parameter in square brackets.
\item a parameter in curly brackets (a \textit{sequence} node in the \gls{mlp-pt}) if none of the above.
\item an $@$ symbols if none of the above.
\item a variable in curly brackets (a \textit{sequence} node) if none of the above.
\end{itemize}
\begin{algorithm}[!ht]
\caption{The translate function of the MacroTranslator. This code ignores error handling.}\label{alg:macro-translation}
	\begin{algorithmic}[1]
	\Require 
		\Statex \textit{macro} - node of the semantic macro. 
		\Statex \textit{args} - list of the following siblings of \textit{macro}. 
		\Statex \textit{lexicon} - lexicon file
	\Ensure 
		\Statex Translated semantic macro.

>>>>

\section{Forward Translations}\label{sec:forward-translation}
An abstract translator class analyzes each node of the parsed tree and delegates them to specialized subtranslators. In this process an object called \gls*{teo} is build. This \gls*{teo} is needed to rebuild a string representation of the mathematical expression after all nodes of the parsed tree were translated.

The parsed tree generated by the \gls*{pom}-tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

The overall forward translation process is explained in figure~\ref{fig:forward-trans}. All translation patterns and related information are stored in the DLMF/DRMF tables. These tables are converted by the \verb|lexicon-creator.jar| to the \verb|DLMF-macros-lexicon.txt| lexicon file. Together with the \verb|global-lexicon.txt| file, the \gls*{pom-pt} will be created by the \gls*{pom}-tagger. The \verb|latex-converter.jar| takes a string representation of a semantic \LaTeX{} expression and uses the \gls*{pom} engine as well as our \verb|Translator| to create an appropriate string representation for a specified \gls*{cas}.

\begin{figure}[ht]
	\vspace{-10pt}
	\centering
	\includegraphics[clip, trim=0.2cm 0.2cm 0.2cm 0.2cm, scale=0.72]{ForwardTranslationProcess.pdf}
	\caption{Process diagram of a forward translation process. The \gls*{pom}-tagger generates the \gls*{pom-pt} based on lexicon and \gls*{json} files. The \gls*{pom-pt} will be translated to different \gls*{cas}.}
	\label{fig:forward-trans}
	\vspace{-10pt}
\end{figure}

\subsection{Analyzing the PoM-Parsed Tree}\label{subsec:analyze-mlp}
Since the \gls*{bnf} does not define rules for semantic macros, each argument of the semantic macro and each $@$ symbol are following siblings of the semantic macro node. That is the reason, why we stored the number of parameters, variables and $@$ symbols in the lexicon files. Otherwise, the translator could not find the end of a semantic macro in the \gls*{pom-pt}.

\begin{figure}[ht]
	\centering
	\includegraphics[clip, trim=0.2cm 0.2cm 0.2cm 0.2cm, scale=0.75]{SyntaxTreeUseCase.pdf}
	\caption{pom-pt for a Jacobi polynomial using the DLMF/DRMF \LaTeX{} macro. Each leaf contains information from the lexicon files.}
	\label{fig:syntax-tree-usecase}
	\vspace{-10pt}
\end{figure}

\begin{wrapfigure}{r}{0.5\textwidth}
\vspace{-20pt}
\begin{minipage}{0.5\textwidth}
\begin{algorithm}[H]
\caption{Simple translation algorithm for the POM-Parse Tree}\label{alg:simple-translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of the parse tree $T$
	\Procedure{translate}{$r$}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$);
	\Else
		\ForAll{children $v_n$ of $r$}
			\State {\scriptsize TRANSLATE}($v_n$);
		\EndFor
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\end{minipage}
\vspace{-14pt}
\end{wrapfigure}

Figure~\ref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from table~\ref{tab:JacobiP-usecase}. Because of the differences to expression trees, a backward conversion of the \gls*{pom-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls*{teo}.

With these tools, we can translate a \LaTeX{} expression by translating the \gls*{pom-pt} node by node and perform group or reordering operations for some special cases. The algorithm is realized in a simple recursive structure. The algorithm~\ref{alg:simple-translation} explains the translation process. Whenever the algorithm finds a leaf, it can translate this single term. If the node is not a leaf, it starts to translate all children of the node recursively.

\begin{algorithm}[H]
\caption{Abstract translation algorithm to translate MLP-Parse trees.}\label{alg:translation}
	\begin{algorithmic}[1]
	\Require Root $r$ of a POM-Parse tree $T$. List \textit{following\_siblings} with the following siblings of $r$. The list can be empty.
	\Procedure{abstract\_translator}{$r$, \textit{following\_siblings}}
	\If{$r$ is leaf}
		\State {\scriptsize TRANSLATE\_LEAF}($r$, \textit{following\_siblings});\label{line:transLeaf}
	\Else
		\State \textit{siblings} $ = r$.getChildren(); \Comment{\textit{siblings} is a list of children}
		\State {\scriptsize ABSTRACT\_TRANSLATOR}(\textit{siblings}.removeFirst(), \textit{siblings});\label{line:transSeq}
	\EndIf
	\If{\textit{following\_siblings} is not empty}\label{line:ifNotEmpty}
		\State $r =$ \textit{following\_siblings}.removeFirst();
		\State {\scriptsize ABSTRACT\_TRANSLATOR}($r$, \textit{following\_siblings});
	\EndIf
	\EndProcedure
	\end{algorithmic}
\end{algorithm}
\vspace{-20px}

This approach is not completely feasible, since the algorithm needs to look ahead and check the following siblings in some cases, e.g., in the case of a semantic macro with arguments (see figure~\ref{fig:syntax-tree-usecase}). The {\footnotesize \verb|TRANSLATE_LEAF|} function needs to know the following siblings of the current node. Algorithm~\ref{alg:translation} explains a more detailed but abstract version of the final algorithm of the presented translation tool.

If the root $r$ is a leaf, it still can be translated as a leaf. Eventually, some of the following siblings are needed in order to translate $r$. The list of \textit{following\_siblings} in line~\ref{line:transLeaf} might be reduced to avoid multiple translations for one node. If $r$ is not a leaf, it contains one or more children. Therefore, we can call the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively for the children. Once we have translated $r$, we can go a step further and translate the next node. Line~\ref{line:ifNotEmpty} checks if there are following siblings left and calls the {\footnotesize \verb|ABSTRACT_TRANSLATOR|} recursively in that case. Translated expressions are stored by the \gls*{teo} object. The detailed translation process is not included in the algorithm.

Algorithm~\ref{alg:translation} is a simplified version of the translator process. The main progress happens in \cref{line:transLeaf,line:transSeq}. There are several cases what $r$ can be. Table~\ref{tab:allTypesTable} on page \pageref{tab:allTypesTable} gives an overview of all different types in a \gls*{pom-pt}. A more detailed explanation of the types can be found in~\parencite{POM-Tagger}.

The \gls*{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols, similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegates the translation to specialized subtranslators. This delegation process is implemented in \cref{line:transLeaf,line:transSeq} of algorithm~\ref{alg:translation}. The Subsection~\ref{sec:subtranslators} discusses these classes in more detail.

This approach seems to work for most of semantic \LaTeX{} expressions. However, there is one problem with this straight forward approach which was mentioned shortly mentioned in section~\ref{sec:problems}. There are many different types of notations used to represent formulae. Table~\ref{tab:notations} illustrates the simple expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between their operands, which only makes sense as long as the operator is a binary operator.

\begin{wrapfigure}{r}{0.4\textwidth}
\vspace{-22pt}
\begin{minipage}{0.4\textwidth}
\center
\begin{tabular}{lc}
	\hline
	Notation & Expression \\
	\hline
	Infix & $(a+b) \cdot x$\\
	Prefix & $\cdot + a\ b\ x$\\
	Postfix & $a\ b + x\ \cdot$\\
	Functional & $\cdot(+(a, b), x)$\\
	\hline
\end{tabular}
\vspace{-5pt}
\caption{The mathematical expression '$(a+b) \cdot x$' in infix, prefix, postfix and functional notation.}
\label{tab:notations}
\end{minipage}
\vspace{-15pt}
\end{wrapfigure}

In mathematical expressions, notations are mostly mixed, depending on the case and number of operands. For example, infix notation is common for binary operators ($+$, $-$, $\cdot$, $\mod$ etc.), while functional notations are conveniently used for any kind of functions ($sin$, $cos$, etc.). Sometimes the same symbol is used in different notations to distinguish different meanings. For example, the '$-$' as an unary operator is used in prefix notation to indicate the negative value of its operand, such as in '$-2$'. Of course, $-$ can also be the binary operator for subtraction, which is commonly used in infix notation. An example for the postfix notation is factorial, such as '$2!$'.

\input{chapters/03-translator/all-types-table.tex}

Most programming languages (and \gls*{cas} as well) internally use prefix or postfix notation and do not mix the notations in one expression, since it more convenient to parse expressions in uniform notations. However, the common practice in science is to use mixed notations in expressions. Since the \gls*{pom} has rarely implemented mathematical grammatical rules yet, it takes the input as it is and does not build an expression tree. Therefore, it parses all four examples from table~\ref{tab:notations} to four different \gls*{pom-pt}s rather than to one unique expression tree. In general, this is not a problem for our translation process since most \gls*{cas} are familiar with the most common notations. Therefore, the translator does not need to know that $a$ and $b$ are the operands of the binary operator '$+$' in '$a+b$'. The translator could simply translates the symbols in '$a+b$' in the same order as they appear in the expression and the \gls*{cas} would understand it. However, this simple approach generates two problems.
\begin{enumerate}%[label=\arabic*)]
\item \label{prob:1} The translated expression is only syntactically correct if the input expression was syntactically correct.
\item \label{prob:2} We cannot translate expressions to a \gls*{cas} which uses a non-standard notation.
\end{enumerate}

Problem~\ref{prob:1} should be obvious. Since we want to develop a translation tool and not a verification tool for mathematical \LaTeX{} expressions, we can assume syntactically correct input expressions and produce errors otherwise. Problem~\ref{prob:2} is more difficult to solve. If a user wants to support a \gls*{cas} that uses prefix or postfix notation by default, the translator would fail in its current state. Supporting \gls*{cas} with another notations would be a part of future work.

@@

In most cases, parentheses can be used to disambiguate expressions. We used them in table~\ref{tab:amb_ex} to clarify the different interpretations in example 2. But sometimes, even parentheses cannot solve a mistaken computation. For example, there is no way to add parentheses to force \Maple{} to compute $n!!$ as the double factorial function. Even $(n)!!$ will be interpreted as $(n!)!$. Rather than using the exclamation mark in \Maple, one could also use the functional notation. For example, the interpretations $(2!)!$ and $(2)!!$ can be distinguished in \Maple{} by using \verb|factorial(factorial(2))| and \verb|doublefactorial(2)| respectively. We define the translations as follows:

\begin{eqnarray*}
\verb|n!| &\overset{\langMaple}{\mapsto}& \verb|factorial(n)|,\\
\verb|n!!| &\overset{\langMaple}{\mapsto}& \verb|doublefactorial(n)|.
\end{eqnarray*}
%The translator needs to presume some properties for the functions similar to \LaTeX, which only recognizes the very next element right after the caret for the superscript.

Algorithm~\ref{alg:translation} does not allow this translation right now. It has no access to previously translated nodes in its current state. This problem is solved by the \gls*{teo} that stores and groups translated objects as lists. This allows to access the latest translated expression and use it as the argument for the factorial function. Table~\ref{tab:teo-list} shows three example for the \gls*{teo} list that groups some tokens.

%The biggest problem about translating the factorial (or double factorial) function is the common postfix notation for them. In this stage, algorithm~\ref{alg:translation} only translates the current node and if necessary the following siblings, but not the preceding siblings (see algorithm~\ref{alg:translation}). Consider an expression such as '$4!$'. When the variable node $r$ reaches the exclamation mark, the argument '$4$' of the function has already been translated and is no longer accessible for the current algorithm. This problem occurs whenever a postfix notation is involved. The following two approaches would solve the problem for the problem.

%\begin{enumerate}
%\item\label{appr:1} The translator always checks if the next two following siblings are one or two exclamation marks.
%\item\label{appr:2} The current state of program has access to the previously translated expressions and can modify them retrospectively.
%\end{enumerate}

%Approach~\ref{appr:1} would be unnecessarily inefficient and does not scale well if more of these cases occur. Therefore we implement approach~\ref{appr:2}. That is another reason why we implement an extra object to organize translated expressions, the \gls*{teo}. This object stores all parts of the previously translated expressions. The translator has access to these parts and is able to modify them afterwards.

%\input{chapters/03-translator/all-types-table.tex}

%\subsection{Delegation of Translations}\label{subsec:delegation}

%\begin{wrapfigure}{l}{0.45\textwidth}
%	\vspace{-20pt}
%	\centering
%	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.4]{ForwardTranslator.pdf}
%	\caption{A scheme of the forward translator and its specialized subtranslators.}
%	\label{fig:ForwardDel}
%	\vspace{-25pt}
%\end{wrapfigure}

%Algorithm~\ref{alg:translation} is a simplified version of the translator process. The main progress happens in \cref{line:transLeaf,line:transSeq}. There are several cases what $r$ can be. Table~\ref{tab:allTypesTable} on page \pageref{tab:allTypesTable} gives an overview of all different types in a \gls*{pom-pt}. A more detailed explanation of the types can be found in~\parencite{POM-Tagger}.

%The \gls*{bnf} grammar defines some basic grammatical rules for generic \LaTeX{} macros, such as for \verb|\frac|, \verb|\sqrt|. Therefore, there is a hierarchical structure for those symbols, similar to the structure in expression trees. As already mentioned, some of these types can be translated directly, such as Greek letters, while others are more complex, such as semantic \LaTeX{} macros. Therefore, the translators delegates the translation to specialized subtranslators. This delegation process is implemented in \cref{line:transLeaf,line:transSeq} of algorithm~\ref{alg:translation}.

%Figure~\ref{fig:ForwardDel} is a scheme of the forward translator and its subtranslators. The entry point for each translation process of the program is the \verb|SemanticLatexTranslator|. It delegates the translations for the expression through the \verb|AbstractTranslator| to the other subtranslators. The \verb|AbstractTranslator| is the interface of all other translators and coordinates the translation process by delegating translations for specific nodes to specialized classes. For example, the \verb|MacroTranslator| only translates semantic \LaTeX{} macros. The \verb|EmptyExpressionTranslator| is used to delegate the so called \textit{empty expressions} to other classes. An \textit{empty expression} is a node with children, but without a symbol itself, such as \textit{sequences} and \textit{balanced expressions}.

\subsection{Subtranslators}\label{sec:subtranslators}
%The most interesting translators are the \verb|MacroTranslator| (translating semantic \LaTeX{} macros) and the \verb|SequenceTranslator| (translating all kinds of \textit{sequences}).

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balances of parentheses in expressions. An expression such as '$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as '$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. In addition, the parenthesis checker could simply be deactivated to allow mismatched parentheses in an expression. But this functionality is not implemented yet.

The \verb|SequenceTranslator| also handle positions of multiplication symbols. There are a couple of obvious choices to translate multiplication signs. The most common symbol for multiplications is still the white space or none symbol at all, as explained previously. Consider the simple expression $2n\pi$. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as $f(x)$. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.

%Besides the translation of sequences, the translation process of the \Macro s can be complex too; not only because of the structure of the \gls*{pom-pt}, but also because of the complex definition of the semantic macros. As explained in \ref{subsec:macros}, they can contain optional parameters, exponents are allowed before and after the arguments and the number of $@$ symbols varies. All these different cases will be solved by the \verb|MacroTranslator|.

\begin{algorithm}[!ht]
\caption{The translate function of the MacroTranslator. This code ignores error handling.}\label{alg:macro-translation}
	\begin{algorithmic}[1]
	\Require 
		\Statex \textit{macro} - node of the semantic macro. 
		\Statex \textit{args} - list of the following siblings of \textit{macro}. 
		\Statex \textit{lexicon} - lexicon file
	\Ensure 
		\Statex Translated semantic macro.
